Approximation as Differentiation

### 2025-05-07 main/formal-structures.tex
LatGal.

\noindent The $\adj$ notation is justified because a Galois connection $f \adj g: X \to Y$ can also be seen an
adjunction between poset-as-categories, with monotone $f$ and $g$ interpreted as functors; $f$ is usually
referred to as the \emph{upper} (right) adjoint and $g$ as the \emph{lower} (left) adjoint. Galois connections
compose component-wise, with $\id_X \adj \id_X: X \to X$ as the unit for composition. Thus we can define the
following category.

@@

It is easy to verify that $+$ is associative and commutative and that $f + 0 = f = 0 + f$, and so every
hom-object $\cat{C}(X,Y)$ is an object in $\CMon$. Moreover composition is \emph{bilinear}, i.e.~given by a
family of morphisms $\Hom{\cat{C}}{Y}{Z} \tensor \Hom{\cat{C}}{X}{Y} \to \Hom{\cat{C}}{X}{Z}$ in $\CMon$ that
preserve the additive structure in $\Hom{\cat{C}}{Y}{Z}$ and $\Hom{\cat{C}}{X}{Y}$ separately:

@@

\note{Probably a better way: (1) introduce the category of lattices and Galois connections, which is what we will use to interpret forward and backward approximation, and establish some properties of this category; (2) say that we want sets with each point having an associated lattice of approximations, and morphisms to match; (3) note that this is exactly the Grothendieck construction for a particular indexed category; (4) Say that CHAD is what we need here; (5) Recapitulate the bits of CHAD that we need, with our Agda formalisation; (6) Explain how we cope with higher-order structure, by using presheaves and the Yoneda embedding.}

>>>>

\noindent The $\adj$ notation is justified because a Galois connection $f \adj g: X \to Y$ can also be seen an
adjunction between poset-as-categories, with monotone $f$ and $g$ interpreted as functors; $f$ is usually
referred to as the \emph{upper} (right) adjoint and $g$ as the \emph{lower} (left) adjoint. Galois connections
compose component-wise, with $\id_X \adj \id_X: X \to X$ as the unit for composition, and thus form a category
where the objects are posets and the morphisms are Galois connections.

As indicated in \secref{introduction:galois-slicing}, for Galois slicing we require a setting where the
approximants of a point form a bounded lattice, with least element $\bot$ representing the approximation that
discards all information and greatest element $\top$ representing the approximation that discards no
information; so we define the following subcategory.

\begin{definition}
Define the category $\LatGal$ which has as objects $X, Y$ all bounded lattices, and as morphisms from $X$ to
$Y$, all Galois connections $f \adj g: X \to Y$.
\end{definition}

\noindent Right adjoints preserves limits and left adjoints preserves colimits, so for any $f \adj g: X \to Y$
in $\LatGal$, $f$ and $g$ preserve the meet-semilattice structure of $X$ and join-semilattice structure of $Y$
respectively.

@@

It is easy to verify that $+$ is associative and commutative and that $f + 0 = f = 0 + f$, and thus that every
hom-object $\cat{C}(X,Y)$ is an object in $\CMon$. Moreover composition is \emph{bilinear}, i.e.~given by a
family of morphisms $\Hom{\cat{C}}{Y}{Z} \tensor \Hom{\cat{C}}{X}{Y} \to \Hom{\cat{C}}{X}{Z}$ in $\CMon$ that
preserve the additive structure in $\Hom{\cat{C}}{Y}{Z}$ and $\Hom{\cat{C}}{X}{Y}$ separately:

@@

% Done:
% (1) introduce the category of lattices and Galois connections, which is what we will use to interpret forward and backward approximation, and establish some properties of this category;

\note{Probably a better way: (2) say that we want sets with each point having an associated lattice of approximations, and morphisms to match; (3) note that this is exactly the Grothendieck construction for a particular indexed category; (4) Say that CHAD is what we need here; (5) Recapitulate the bits of CHAD that we need, with our Agda formalisation; (6) Explain how we cope with higher-order structure, by using presheaves and the Yoneda embedding.}
